---
title: "Stat 184 Free Throw Analysis"
author: "Robert Dang, Connor Box, Aidan Wieners"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
execute:
  echo: false
  warning: false
  message: false
---

# Introduction

This report presents an exploratory data analysis of NBA player attributes and their relationship to free throw percentage among players with at least 150 free throw attempts. The goal is to investigate whether factors such as player role, physical characteristics, and playing time are associated with free throw shooting performance.

---

# Data Provenance

---

# FAIR and CARE Principles

---

# Data Cleaning and Feature Creation

Before analysis, the dataset was cleaned and transformed. This included:

- Removing percentage symbols from free throw percentage
- Removing commas from total minutes
- Converting variables to numeric format

These transformations ensure that the data are consistent and suitable for analysis.

---

# Exploratory Data Analysis

## Robert

---

## Connor

This section examines whether a player’s height is related to their free throw percentage within each position. To evaluate this, we calculate the correlation between height and free throw percentage for each position.

```{r}
library(tidyverse)
library(knitr)

nba_raw <- read_csv("NBAPlayerStats.csv")

nba_clean <- nba_raw |>
  rename(
    position = Position,
    ft_percent = `FT%`,
    height_in = `Height (in)`
  ) |>
  mutate(
    ft_percent = as.numeric(str_remove(ft_percent, "%")),
    height_in = as.numeric(height_in)
  )

position_corr <- nba_clean |>
  group_by(position) |>
  summarise(
    correlation = round(cor(height_in, ft_percent, use = "complete.obs"), 2),
    avg_ft = round(mean(ft_percent, na.rm = TRUE), 1),
    avg_height = round(mean(height_in, na.rm = TRUE), 1),
    n_players = n(),
    .groups = "drop"
  )
```

As shown in @fig-connor, most positions exhibit weak relationships between height and free throw percentage.

```{r}
#| label: fig-connor
#| fig-cap: "Correlation between height and free throw percentage by position."
#| fig-alt: "Bar chart showing correlation between height and free throw percentage for each NBA position."
#| fig-pos: "H"

ggplot(position_corr, aes(x = position, y = correlation)) +
  geom_col(fill = "skyblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Position",
    y = "Correlation"
  ) +
  theme_minimal()
```

@tbl-connor provides the corresponding numerical values.

```{r}
#| label: tbl-connor
#| tbl-cap: "Correlation between height and free throw percentage by position."

kable(
  position_corr,
  col.names = c("Position", "Correlation", "Avg FT%", "Avg Height", "Players")
)
```

The results suggest that height is not a strong predictor of free throw performance, as correlations remain small across most positions.

---

## Aidan

---

# Overall Data Story

---

# Conclusion

---

# Open Science and Reproducibility

---

# Author Contributions

- Robert Dang:
- Connor Box:
- Aidan Wieners:

---

# Appendix: Code

```{r}
#| eval: false
#| echo: true

library(tidyverse)
library(knitr)

nba_raw <- read_csv("NBAPlayerStats.csv")

nba_clean <- nba_raw |>
  rename(
    position = Position,
    ft_percent = `FT%`,
    height_in = `Height (in)`
  ) |>
  mutate(
    ft_percent = as.numeric(str_remove(ft_percent, "%")),
    height_in = as.numeric(height_in)
  )

position_corr <- nba_clean |>
  group_by(position) |>
  summarise(
    correlation = round(cor(height_in, ft_percent, use = "complete.obs"), 2),
    avg_ft = round(mean(ft_percent, na.rm = TRUE), 1),
    avg_height = round(mean(height_in, na.rm = TRUE), 1),
    n_players = n(),
    .groups = "drop"
  )

ggplot(position_corr, aes(x = position, y = correlation)) +
  geom_col(fill = "skyblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Position",
    y = "Correlation"
  ) +
  theme_minimal()

kable(
  position_corr,
  col.names = c("Position", "Correlation", "Avg FT%", "Avg Height", "Players")
)
```