---
title: "Stat 184 Free Throw Analysis"
author: "Robert Dang, Connor Box, Aidan Wieners"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
execute:
  echo: false
  warning: false
  message: false
---

# Introduction

This report presents a data analysis and exploration of the question: "What makes a good NBA free throw shooter?" We used player height and position and examined their relationship to free throw percentage among players with at least 150 free throw attempts during the 2025-2026 regular season.


# Data Provenance

## Who Collected It
The data was manually collected and compiled by our team. Player 
statistics were gathered by hand from Databallr (databallr.com) and 
entered into an Excel spreadsheet, which was then saved as a CSV file 
for use in this analysis.

## What the Data Is
Each case (row) in the dataset represents one NBA player during the 
2025-2026 regular season. The attributes include player name, team, 
position, physical measurements (height and wingspan in inches), and 
performance statistics such as free throw percentage (FT%) and free 
throw attempts (FTA).

## When It Was Collected
The data was manually recorded on 4/28/2026. The  
statistics reflect the 2025-2026 NBA regular season as displayed 
on Databallr at the time of collection.

## Where It Came From
Statistics were transcribed by hand from databallr.com into an Excel 
spreadsheet. The original source of the underlying data is the NBA's 
official game and player records.

## Why It Was Collected
The data was collected to investigate how player height and position relate to free 
throw shooting percentage.

## How It Was Collected
Team members used the features on Databallr to retrieve the necessary statistics. manually looked up each player on Databallr and recorded into an Excel file. This file was then exported as a CSV for use in R. 

# FAIR Principles
- **Findable:** The original statistics are publicly viewable on 
  databallr.com. Our compiled dataset is stored in our team's 
  GitHub repository.
- **Accessible:** The CSV file is available in our repo and can be 
  downloaded by anyone with access to it.
- **Interoperable:** The CSV format is compatible with many coding languages
- **Reusable:** Column names are clearly labeled and the data 
  context is well documented in this report, making it reusable 
  for future analysis.

# CARE Principles
This dataset covers NBA players whose statistics are part 
of the public record. No private or sensitive personal information 
was collected beyond what is officially published by the league. 

# Data Cleaning and Feature Creation

Before analysis, the dataset was cleaned and transformed. This included:

- Removing percentage symbols from free throw percentage
- Removing commas from total minutes
- Converting variables to numeric format

These transformations ensure that the data are consistent and suitable for analysis.


# Exploratory Data Analysis

## Robert

In this section, we look at if the height and position of the players and see if they correlate, which is the basis of our overall analysis. 

```{r}
#| echo: false
#| warning: false
#| message: false

# Tidyverse Style Guide

# Step 1: Load packages
library(ggplot2)
library(dplyr)
library(readr)
library(ggrepel)
library(knitr)
library(kableExtra)

# Step 2: Load and tidy the data
df <- read_csv("NBAPlayerStats.csv") |>
  rename(
    ft_pct = `FT%`,
    height = `Height (in)`
  ) |>
  mutate(
    ft_pct = as.numeric(gsub("%", "", ft_pct))
  )

# Step 3: Summarize average height and FT% per position
position_summary <- df |>
  group_by(Position) |>
  summarise(
    avg_height = mean(height, na.rm = TRUE),
    avg_ft_pct = mean(ft_pct, na.rm = TRUE),
    count      = n()
  )

# Plot for height distribution by position
df |>
  mutate(Position = reorder(Position, height)) |>
  ggplot(aes(x = Position, y = height, fill = Position)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.size = 2) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = function(x) paste0(x, '"')) +
  labs(
    title    = "Height Distribution by Position",
    subtitle = "Positions ordered by median height",
    x        = "Position",
    y        = "Height (inches)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "grey40"),
    legend.position = "none"
  )

# Table for height summary by position
df |>
  group_by(Position) |>
  summarise(
    `Player Count`       = n(),
    `Avg Height (in)`    = round(mean(height, na.rm = TRUE), 1),
    `Median Height (in)` = round(median(height, na.rm = TRUE), 1),
    `Min Height (in)`    = round(min(height, na.rm = TRUE), 1),
    `Max Height (in)`    = round(max(height, na.rm = TRUE), 1)
  ) |>
  arrange(desc(`Avg Height (in)`)) |>
  kable(
    caption = "Height Summary by Position for Players Who Shot At Least 150 FT's (tallest to shortest)",
    align   = c("l", "c", "c", "c", "c", "c")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE
  )
```

The box plot and summary table together confirm a clear and consistent 
relationship between position and height in the 2025-2026 NBA season. 
From the table, we see that centers are the tallest position with an average height of 82.9 inches, 
while point guards are the shortest at an average of 74.6 inches, a 
difference of over 8 inches. 

Looking at the box plot reveals that the positions 
follow a staircase pattern from PG to C, suggesting that 
height is a strong defining characteristic of positional assignment in 
the NBA. Also, most positions have relatively tight height 
distributions, meaning players at the same position tend to be similar 
in height. Centers show the widest spread, ranging from 80.8 to 88.0 
inches, indicating more variation at that position than any other. 
Point guards, by contrast, are tightly clustered between 72.5 and 77.5 
inches, with very little variation.


## Connor

This section examines whether a player’s height is related to their free throw percentage within each NBA position. While Robert showed that height differs significantly across positions, this analysis focuses on whether taller or shorter players within the same role tend to shoot free throws more effectively. To evaluate this relationship, we calculate the correlation between height and free throw percentage for each position.

```{r}
library(tidyverse)
library(knitr)

nba_raw <- read_csv("NBAPlayerStats.csv")

nba_clean <- nba_raw |>
  rename(
    position = Position,
    ft_percent = `FT%`,
    height_in = `Height (in)`
  ) |>
  mutate(
    ft_percent = as.numeric(str_remove(ft_percent, "%")),
    height_in = as.numeric(height_in)
  )

position_corr <- nba_clean |>
  group_by(position) |>
  summarise(
    correlation = round(cor(height_in, ft_percent, use = "complete.obs"), 2),
    avg_ft = round(mean(ft_percent, na.rm = TRUE), 1),
    avg_height = round(mean(height_in, na.rm = TRUE), 1),
    n_players = n(),
    .groups = "drop"
  )
```

As shown in @fig-connor, most positions exhibit correlations close to zero, indicating little to no linear relationship between height and free throw percentage within those groups. This suggests that among players who already share a similar role and body type, being slightly taller or shorter does not meaningfully impact free throw shooting ability.

```{r}
#| label: fig-connor
#| fig-cap: "Correlation between height and free throw percentage by position."
#| fig-alt: "Bar chart comparing the correlation between player height and free throw percentage across NBA positions (C, PF, PG, SF, SG). Most correlations are close to zero, indicating little relationship between height and free throw shooting within positions. Centers show a small positive correlation, while shooting guards show the strongest negative correlation."
#| fig-pos: "H"

ggplot(position_corr, aes(x = position, y = correlation)) +
  geom_col(fill = "skyblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Position",
    y = "Correlation"
  ) +
  theme_minimal()
```

@tbl-connor provides the corresponding numerical values. Centers show a modest positive correlation (0.27), while shooting guards show a more noticeable negative correlation (-0.47). However, because the results vary across positions and most correlations remain relatively weak, the overall pattern suggests that height alone is not a consistent predictor of free throw performance within a position group.

```{r}
#| label: tbl-connor
#| tbl-cap: "Correlation between height and free throw percentage by position."
#| tbl-pos: "H"

kable(
  position_corr,
  col.names = c("Position", "Correlation", "Avg FT%", "Avg Height", "Players")
)
```

These results complement Robert’s findings by showing that although height is strongly associated with NBA position, it does not explain much of the variation in free throw percentage once players are compared within the same position. This also helps support Aidan’s broader positional analysis, which examines how average free throw percentages differ across positions overall. Together, the analyses suggest that free throw shooting ability is driven more by individual skill and technique than by small differences in player height.



## Aidan
Robert confirmed that height varies by position and Connor showed that height has little correlation with accuracy within each position, so I am zooming out and looking at the broader question: 
When we look at each position as a whole, which position tends to shoot free throws better on average?
To explore this, I summarized the data by position and created a bubble chart where each bubble represents one position to show the correlation.

```{r}
#| echo: false

#Tidyverse style guide
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)

# Tidy the data
nba <- read_csv("NBAPlayerStats.csv") |>
  rename(
    ft_pct   = `FT%`,
    height   = `Height (in)`,
    position = Position
  ) |>
  mutate(
    ft_pct = as.numeric(gsub("%", "", ft_pct)),
    height = as.numeric(height)
  )

# Summarize by position
pos_summary <- nba |>
  group_by(position) |>
  summarise(
    avg_height = round(mean(height, na.rm = TRUE), 1),
    avg_ft_pct = round(mean(ft_pct,  na.rm = TRUE), 1),
    n_players  = n(),
    .groups = "drop"
  )
```
@plot-aidan makes the trend pretty easy to see. You can see how there is little difference in the average FT% in guards and small forwards, who also tend to shoot much more accurate than bigs, as they're all fairly even on the Y-axis. The scatter plot helps show a pretty consistent drop off in FT% as height rises, shown by about a 5% drop from small forwards to power forwards, and another 5% drop from power forwards to centers.
We can see that players who are either point guards, shooting guards, or small forwards are on average 79 inches or shorter. From this we can make the observation that players in those positions typically have the ideal attributes to shoot a much more accurate FT% compared to the larger positions in power forwards and centers.  

```{r}
#Plot-aidan — Average height vs. FT% by position
#| label: plot-aidan
#| fig-cap: "Average height vs. average FT% by position. Bubble size shows number of players in each group."

ggplot(pos_summary, aes(x = avg_height, y = avg_ft_pct,
                        size = n_players, color = position)) +
  geom_point(alpha = 0.8) +
  geom_text(aes(label = position), color = "black", size = 4, hjust = 0.5, vjust = 0.5)+  
  scale_size_continuous(range = c(6, 14), name = "# of Players") +
  scale_color_brewer(palette = "Set1", guide = "none") +
  labs(
    title = "Average Height vs. Average Free Throw % by Position",
    x = "Average Height (inches)",
    y = "Average Free Throw %"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold")
  )
``` 
@table-aidan helps viewers clearly see how close the FT% really is between point guards, shooting guards, and small forwards, even given the signifigant average height difference from PG-SF.
Having the actual numbers displayed in addition to the table really helps add emphasis on those three positions shooting within 1.1% of each other on average. 

```{r}
# Table-aidan — Average height vs. FT% by position

#| label: tab-aidan
#| fig-cap: "Provides the corresponding numerical values (Position, Avg Height, Avg FT%, # of Players)."
pos_summary |>
  arrange(desc(avg_ft_pct)) |>
  kable(
    caption = "Average Height and Free Throw Percentage by Position",
    col.names = c("Position", "Avg Height (in)", "Avg Free Throw %", "# Players"),
    align = c("l", "c", "c", "c")
  )
```

# Overall Conclusion


# Open Science and Reproducibility


# Author Contributions

- Robert Dang:
- Connor Box:
- Aidan Wieners:


# AI Usage 


# Appendix: Code
Robert's code 
```{r}
#| eval: false
#| echo: true
#| results: hide

# Tidyverse Style Guide

# Step 1: Load packages
library(ggplot2)
library(dplyr)
library(readr)
library(ggrepel)
library(knitr)
library(kableExtra)

# Step 2: Load and tidy the data
df <- read_csv("NBAPlayerStats.csv") |>
  rename(
    ft_pct = `FT%`,
    height = `Height (in)`
  ) |>
  mutate(
    ft_pct = as.numeric(gsub("%", "", ft_pct))
  )

# Step 3: Summarize average height and FT% per position
position_summary <- df |>
  group_by(Position) |>
  summarise(
    avg_height = mean(height, na.rm = TRUE),
    avg_ft_pct = mean(ft_pct, na.rm = TRUE),
    count      = n()
  )

# Plot for height distribution by position
df |>
  mutate(Position = reorder(Position, height)) |>
  ggplot(aes(x = Position, y = height, fill = Position)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.size = 2) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = function(x) paste0(x, '"')) +
  labs(
    title    = "Height Distribution by Position",
    subtitle = "Positions ordered by median height",
    x        = "Position",
    y        = "Height (inches)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "grey40"),
    legend.position = "none"
  )

# Table for height summary by position
df |>
  group_by(Position) |>
  summarise(
    `Player Count`       = n(),
    `Avg Height (in)`    = round(mean(height, na.rm = TRUE), 1),
    `Median Height (in)` = round(median(height, na.rm = TRUE), 1),
    `Min Height (in)`    = round(min(height, na.rm = TRUE), 1),
    `Max Height (in)`    = round(max(height, na.rm = TRUE), 1)
  ) |>
  arrange(desc(`Avg Height (in)`)) |>
  kable(
    caption = "Height Summary by Position for Players Who Shot At Least 150 FT's (tallest to shortest)",
    align   = c("l", "c", "c", "c", "c", "c")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE
  )
```

Connor's code 
```{r}
#| eval: false
#| echo: true


### Author: Connor Box

# Step 1: Load packages
library(tidyverse)
library(knitr)

# Step 2: Load the NBA player data
nba_raw <- read_csv("NBAPlayerStats.csv")

# Step 3: Clean the data and rename columns
nba_clean <- nba_raw |>
  rename(
    position = Position,
    ft_percent = `FT%`,
    height_in = `Height (in)`
  ) |>
  mutate(
    ft_percent = as.numeric(str_remove(ft_percent, "%")),
    height_in = as.numeric(height_in)
  )

# Step 4: Calculate the correlation between height and free throw percentage by position
position_corr <- nba_clean |>
  group_by(position) |>
  summarise(
    correlation = round(cor(height_in, ft_percent, use = "complete.obs"), 2),
    avg_ft = round(mean(ft_percent, na.rm = TRUE), 1),
    avg_height = round(mean(height_in, na.rm = TRUE), 1),
    n_players = n(),
    .groups = "drop"
  )

# Step 5: Create a bar chart showing the correlations by position
#| label: fig-connor
#| fig-cap: "Correlation between height and free throw percentage by position."
#| fig-alt: "Bar chart comparing the correlation between player height and free throw percentage across NBA positions (C, PF, PG, SF, SG). Most correlations are close to zero, indicating little relationship between height and free throw shooting within positions. Centers show a small positive correlation, while shooting guards show the strongest negative correlation."
#| fig-pos: "H"

ggplot(position_corr, aes(x = position, y = correlation)) +
  geom_col(fill = "skyblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Position",
    y = "Correlation"
  ) +
  theme_minimal()

# Step 6: Create a table showing the correlation values and summary statistics
#| label: tbl-connor
#| tbl-cap: "Correlation between height and free throw percentage by position."
#| tbl-pos: "H"

kable(
  position_corr,
  col.names = c("Position", "Correlation", "Avg FT%", "Avg Height", "Players")
)
```
Aidan's Code
```{r}
#| eval: false
#| echo: true

### Author: Aidan Wieners
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)

# Tidy the data
nba <- read_csv("NBAPlayerStats.csv") |>
  rename(
    ft_pct   = `FT%`,
    height   = `Height (in)`,
    position = Position
  ) |>
  mutate(
    ft_pct = as.numeric(gsub("%", "", ft_pct)),
    height = as.numeric(height)
  )

# Summarize by position
pos_summary <- nba |>
  group_by(position) |>
  summarise(
    avg_height = round(mean(height, na.rm = TRUE), 1),
    avg_ft_pct = round(mean(ft_pct,  na.rm = TRUE), 1),
    n_players  = n(),
    .groups = "drop"
  )

#Plot-aidan — Average height vs. FT% by position
#| label: plot-aidan
#| fig-cap: "Average height vs. average FT% by position. Bubble size shows number of players in each group."

ggplot(pos_summary, aes(x = avg_height, y = avg_ft_pct,
                        size = n_players, color = position)) +
  geom_point(alpha = 0.8) +
  geom_text(aes(label = position), color = "black", size = 4, hjust = 0.5, vjust = 0.5)+  
  scale_size_continuous(range = c(6, 14), name = "# of Players") +
  scale_color_brewer(palette = "Set1", guide = "none") +
  labs(
    title = "Average Height vs. Average Free Throw % by Position",
    x = "Average Height (inches)",
    y = "Average Free Throw %"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold")
  )

# Table-aidan — Average height vs. FT% by position

#| label: tab-aidan
#| fig-cap: "Provides the corresponding numerical values (Position, Avg Height, Avg FT%, # of Players)."
pos_summary |>
  arrange(desc(avg_ft_pct)) |>
  kable(
    caption = "Average Height and Free Throw Percentage by Position",
    col.names = c("Position", "Avg Height (in)", "Avg Free Throw %", "# Players"),
    align = c("l", "c", "c", "c")
  )
```
