---
title: "Stat 184 Free Throw Analysis"
author: "Robert Dang, Connor Box, Aidan Wieners"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
execute:
  echo: false
  warning: false
  message: false
---

# Introduction

This report presents a data analysis and exploration of the question: "What makes a good NBA free throw shooter?" We used player height and position and examined their relationship to free throw percentage among players with at least 150 free throw attempts during the 2025-2026 regular season.


# Data Provenance

## Who Collected It
The data was manually collected and compiled by our team. Player 
statistics were gathered by hand from Databallr (databallr.com) and 
entered into an Excel spreadsheet, which was then saved as a CSV file 
for use in this analysis.

## What the Data Is
Each case (row) in the dataset represents one NBA player during the 
2025-2026 regular season. The attributes include player name, team, 
position, physical measurements (height and wingspan in inches), and 
performance statistics such as free throw percentage (FT%) and free 
throw attempts (FTA).

## When It Was Collected
The data was manually recorded on 4/28/2026. The  
statistics reflect the 2025-2026 NBA regular season as displayed 
on Databallr at the time of collection.

## Where It Came From
Statistics were transcribed by hand from databallr.com into an Excel 
spreadsheet. The original source of the underlying data is the NBA's 
official game and player records.

## Why It Was Collected
The data was collected to investigate how player height and position relate to free 
throw shooting percentage.

## How It Was Collected
Team members used the features on Databallr to retrieve the necessary statistics. manually looked up each player on Databallr and recorded into an Excel file. This file was then exported as a CSV for use in R. 

# FAIR Principles
- **Findable:** The original statistics are publicly viewable on 
  databallr.com. Our compiled dataset is stored in our team's 
  GitHub repository.
- **Accessible:** The CSV file is available in our repo and can be 
  downloaded by anyone with access to it.
- **Interoperable:** The CSV format is compatible with many coding languages
- **Reusable:** Column names are clearly labeled and the data 
  context is well documented in this report, making it reusable 
  for future analysis.

# CARE Principles
This dataset covers NBA players whose statistics are part 
of the public record. No private or sensitive personal information 
was collected beyond what is officially published by the league. 

# Data Cleaning and Feature Creation

Before analysis, the dataset was cleaned and transformed. This included:

- Removing percentage symbols from free throw percentage
- Removing commas from total minutes
- Converting variables to numeric format

These transformations ensure that the data are consistent and suitable for analysis.


# Exploratory Data Analysis

## Robert

In this section, we look at if the height and position of the players and see if they correlate, which is the basis of our overall analysis. 

```{r}
#| echo: false
#| warning: false
#| message: false

# Tidyverse Style Guide

# Step 1: Load packages
library(ggplot2)
library(dplyr)
library(readr)
library(ggrepel)
library(knitr)
library(kableExtra)

# Step 2: Load and tidy the data
df <- read_csv("NBAPlayerStats.csv") |>
  rename(
    ft_pct = `FT%`,
    height = `Height (in)`
  ) |>
  mutate(
    ft_pct = as.numeric(gsub("%", "", ft_pct))
  )

# Step 3: Summarize average height and FT% per position
position_summary <- df |>
  group_by(Position) |>
  summarise(
    avg_height = mean(height, na.rm = TRUE),
    avg_ft_pct = mean(ft_pct, na.rm = TRUE),
    count      = n()
  )

# Plot for height distribution by position
df |>
  mutate(Position = reorder(Position, height)) |>
  ggplot(aes(x = Position, y = height, fill = Position)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.size = 2) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = function(x) paste0(x, '"')) +
  labs(
    title    = "Height Distribution by Position",
    subtitle = "Positions ordered by median height",
    x        = "Position",
    y        = "Height (inches)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "grey40"),
    legend.position = "none"
  )

# Table for height summary by position
df |>
  group_by(Position) |>
  summarise(
    `Player Count`       = n(),
    `Avg Height (in)`    = round(mean(height, na.rm = TRUE), 1),
    `Median Height (in)` = round(median(height, na.rm = TRUE), 1),
    `Min Height (in)`    = round(min(height, na.rm = TRUE), 1),
    `Max Height (in)`    = round(max(height, na.rm = TRUE), 1)
  ) |>
  arrange(desc(`Avg Height (in)`)) |>
  kable(
    caption = "Height Summary by Position for Players Who Shot At Least 150 FT's (tallest to shortest)",
    align   = c("l", "c", "c", "c", "c", "c")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE
  )
```

The box plot and summary table together confirm a clear and consistent 
relationship between position and height in the 2025-2026 NBA season. 
From the table, we see that centers are the tallest position with an average height of 82.9 inches, 
while point guards are the shortest at an average of 74.6 inches, a 
difference of over 8 inches. 

Looking at the box plot reveals that the positions 
follow a staircase pattern from PG to C, suggesting that 
height is a strong defining characteristic of positional assignment in 
the NBA. Also, most positions have relatively tight height 
distributions, meaning players at the same position tend to be similar 
in height. Centers show the widest spread, ranging from 80.8 to 88.0 
inches, indicating more variation at that position than any other. 
Point guards, by contrast, are tightly clustered between 72.5 and 77.5 
inches, with very little variation.


## Connor

This section examines whether a player’s height is related to their free throw percentage within each position. To evaluate this, we calculate the correlation between height and free throw percentage for each position.

```{r}
library(tidyverse)
library(knitr)

nba_raw <- read_csv("NBAPlayerStats.csv")

nba_clean <- nba_raw |>
  rename(
    position = Position,
    ft_percent = `FT%`,
    height_in = `Height (in)`
  ) |>
  mutate(
    ft_percent = as.numeric(str_remove(ft_percent, "%")),
    height_in = as.numeric(height_in)
  )

position_corr <- nba_clean |>
  group_by(position) |>
  summarise(
    correlation = round(cor(height_in, ft_percent, use = "complete.obs"), 2),
    avg_ft = round(mean(ft_percent, na.rm = TRUE), 1),
    avg_height = round(mean(height_in, na.rm = TRUE), 1),
    n_players = n(),
    .groups = "drop"
  )
```

As shown in @fig-connor, most positions exhibit correlations close to zero, indicating little to no linear relationship between height and free throw percentage within those groups. This suggests that among players who already share a similar role and general body type, being slightly taller or shorter does not meaningfully impact free throw performance.


```{r}
#| label: fig-connor
#| fig-cap: "Correlation between height and free throw percentage by position."
#| fig-alt: "Bar chart showing correlation between height and free throw percentage for each NBA position."
#| fig-pos: "H"

ggplot(position_corr, aes(x = position, y = correlation)) +
  geom_col(fill = "skyblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Position",
    y = "Correlation"
  ) +
  theme_minimal()
```

@tbl-connor provides the corresponding numerical values. While centers show a modest positive correlation and shooting guards show a more noticeable negative correlation, these patterns are inconsistent across positions and likely reflect variation within the sample rather than a strong underlying trend.


```{r}
#| label: tbl-connor
#| tbl-cap: "Correlation between height and free throw percentage by position."

kable(
  position_corr,
  col.names = c("Position", "Correlation", "Avg FT%", "Avg Height", "Players")
)
```

These results suggest that while height is clearly associated with position, it is not a strong predictor of free throw shooting within positions. In other words, once players are grouped by position, differences in height do not explain much of the variation in free throw percentage, indicating that shooting ability is driven more by individual skill than physical size.



## Aidan
(Rundown of my part)

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)

# Tidy the data
nba <- read_csv("NBAPlayerStats.csv") |>
  rename(
    ft_pct   = `FT%`,
    height   = `Height (in)`,
    position = Position
  ) |>
  mutate(
    ft_pct = as.numeric(gsub("%", "", ft_pct)),
    height = as.numeric(height)
  )

# Summarize by position
pos_summary <- nba |>
  group_by(position) |>
  summarise(
    avg_height = round(mean(height, na.rm = TRUE), 1),
    avg_ft_pct = round(mean(ft_pct,  na.rm = TRUE), 1),
    n_players  = n(),
    .groups = "drop"
  )

#Plot-aidan — Average height vs. FT% by position
#| label: plot-aidan
#| fig-cap: "Average height vs. average FT% by position. Bubble size shows number of players in each group."

ggplot(pos_summary, aes(x = avg_height, y = avg_ft_pct,
                        size = n_players, color = position)) +
  geom_point(alpha = 0.8) +
  geom_text(aes(label = position), color = "black", size = 4, hjust = 0.5, vjust = 0.5)+  
  scale_size_continuous(range = c(6, 14), name = "# of Players") +
  scale_color_brewer(palette = "Set1", guide = "none") +
  labs(
    title = "Average Height vs. Average Free Throw % by Position",
    x = "Average Height (inches)",
    y = "Average Free Throw %"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey40")
  )
``` 
(Plot description)

```{r}
# Table-aidan — Average height vs. FT% by position

#| label: tab-aidan
#| fig-cap: "Provides the corresponding numerical values (Position, Avg Height, Avg FT%, # of Players)."
pos_summary |>
  arrange(desc(avg_ft_pct)) |>
  kable(
    caption = "Average Height and Free Throw Percentage by Position",
    col.names = c("Position", "Avg Height (in)", "Avg Free Throw %", "# Players"),
    align = c("l", "c", "c", "c")
  )
```
This table helps viewers clearly see how close the FT% really is between point guards, shooting guards, and small forwards, even given the signifigant average height difference from PG-SF.
Having the actual numbers displayed in addition to the table really helps add emphasis on those three positions shooting within 1.1% of each other on average. 

# Overall Conclusion


# Open Science and Reproducibility


# Author Contributions

- Robert Dang:
- Connor Box:
- Aidan Wieners:


# AI Usage 


# Appendix: Code
Robert's code 
```{r}
#| eval: false
#| echo: true
#| results: hide

# Tidyverse Style Guide

# Step 1: Load packages
library(ggplot2)
library(dplyr)
library(readr)
library(ggrepel)
library(knitr)
library(kableExtra)

# Step 2: Load and tidy the data
df <- read_csv("NBAPlayerStats.csv") |>
  rename(
    ft_pct = `FT%`,
    height = `Height (in)`
  ) |>
  mutate(
    ft_pct = as.numeric(gsub("%", "", ft_pct))
  )

# Step 3: Summarize average height and FT% per position
position_summary <- df |>
  group_by(Position) |>
  summarise(
    avg_height = mean(height, na.rm = TRUE),
    avg_ft_pct = mean(ft_pct, na.rm = TRUE),
    count      = n()
  )

# Plot for height distribution by position
df |>
  mutate(Position = reorder(Position, height)) |>
  ggplot(aes(x = Position, y = height, fill = Position)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.size = 2) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = function(x) paste0(x, '"')) +
  labs(
    title    = "Height Distribution by Position",
    subtitle = "Positions ordered by median height",
    x        = "Position",
    y        = "Height (inches)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "grey40"),
    legend.position = "none"
  )

# Table for height summary by position
df |>
  group_by(Position) |>
  summarise(
    `Player Count`       = n(),
    `Avg Height (in)`    = round(mean(height, na.rm = TRUE), 1),
    `Median Height (in)` = round(median(height, na.rm = TRUE), 1),
    `Min Height (in)`    = round(min(height, na.rm = TRUE), 1),
    `Max Height (in)`    = round(max(height, na.rm = TRUE), 1)
  ) |>
  arrange(desc(`Avg Height (in)`)) |>
  kable(
    caption = "Height Summary by Position for Players Who Shot At Least 150 FT's (tallest to shortest)",
    align   = c("l", "c", "c", "c", "c", "c")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE
  )
```

Connor's code 
```{r}
#| eval: false
#| echo: true

### Author: Connor Box

library(tidyverse)
library(knitr)

nba_raw <- read_csv("NBAPlayerStats.csv")

nba_clean <- nba_raw |>
  rename(
    position = Position,
    ft_percent = `FT%`,
    height_in = `Height (in)`
  ) |>
  mutate(
    ft_percent = as.numeric(str_remove(ft_percent, "%")),
    height_in = as.numeric(height_in)
  )

position_corr <- nba_clean |>
  group_by(position) |>
  summarise(
    correlation = round(cor(height_in, ft_percent, use = "complete.obs"), 2),
    avg_ft = round(mean(ft_percent, na.rm = TRUE), 1),
    avg_height = round(mean(height_in, na.rm = TRUE), 1),
    n_players = n(),
    .groups = "drop"
  )

ggplot(position_corr, aes(x = position, y = correlation)) +
  geom_col(fill = "skyblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Position",
    y = "Correlation"
  ) +
  theme_minimal()

kable(
  position_corr,
  col.names = c("Position", "Correlation", "Avg FT%", "Avg Height", "Players")
)
```
