---
title: "Stat 184 Free Throw Analysis"
author: "Robert Dang, Connor Box, Aidan Wieners"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
execute:
  echo: false
  warning: false
  message: false
---

# Introduction

This report presents a data analysis and exploration of the question: "What makes a good NBA free throw shooter?" We used player height and position and examined their relationship to free throw percentage among players with at least 150 free throw attempts during the 2025-2026 regular season.


# Data Provenance

## Who Collected It
The data was manually collected and compiled by our team. Player 
statistics were gathered by hand from Databallr (databallr.com) and 
entered into an Excel spreadsheet, which was then saved as a CSV file 
for use in this analysis.

## What the Data Is
Each case (row) in the dataset represents one NBA player during the 
2025-2026 regular season. The attributes include player name, team, 
position, physical measurements (height and wingspan in inches), and 
performance statistics such as free throw percentage (FT%) and free 
throw attempts (FTA).

## When It Was Collected
The data was manually recorded on 4/28/2026. The  
statistics reflect the 2025-2026 NBA regular season as displayed 
on Databallr at the time of collection.

## Where It Came From
Statistics were transcribed by hand from databallr.com into an Excel 
spreadsheet. The original source of the underlying data is the NBA's 
official game and player records.

## Why It Was Collected
The data was collected to investigate how player height and position relate to free 
throw shooting percentage.

## How It Was Collected
Team members used the features on Databallr to retrieve the necessary statistics. manually looked up each player on Databallr and recorded into an Excel file. This file was then exported as a CSV for use in R. 

# FAIR Principles
- **Findable:** The original statistics are publicly viewable on 
  databallr.com. Our compiled dataset is stored in our team's 
  GitHub repository.
- **Accessible:** The CSV file is available in our repo and can be 
  downloaded by anyone with access to it.
- **Interoperable:** The CSV format is compatible with many coding languages
- **Reusable:** Column names are clearly labeled and the data 
  context is well documented in this report, making it reusable 
  for future analysis.

# CARE Principles
This dataset covers NBA players whose statistics are part 
of the public record. No private or sensitive personal information 
was collected beyond what is officially published by the league. 

# Data Cleaning and Feature Creation

Before analysis, the dataset was cleaned and transformed. This included:

- Removing percentage symbols from free throw percentage
- Removing commas from total minutes
- Converting variables to numeric format

These transformations ensure that the data are consistent and suitable for analysis.


# Exploratory Data Analysis

## Robert

In this section, we look at if the height and position of the players and see if they correlate, which is the basis of our overall analysis. 

```{r}
#| label: plot-robert
#| fig-cap: "Height Distribution by Position for NBA Players with at Least 150 Free Throw Attempts (2025-2026 Season)"
#| fig-alt: "A box plot showing NBA player height distribution across five positions ordered by median height from shortest to tallest: PG, SG, SF, PF, and C. Each box uses a distinct Set2 palette color with semi-transparent jittered points overlaid. Point guards cluster tightly near 74 inches, shooting guards near 77, small forwards near 79, and power forwards near 80. Centers have the highest median near 83 inches and the widest spread, with outliers exceeding 87 inches. The staircase pattern across positions suggests height is strongly associated with positional assignment in the NBA.
#| echo: false
#| warning: false
#| message: false

# Tidyverse Style Guide

# Step 1: Load packages
library(ggplot2)
library(dplyr)
library(readr)
library(ggrepel)
library(knitr)
library(kableExtra)

# Step 2: Load and tidy the data
df <- read_csv("NBAPlayerStats.csv") |>
  rename(
    ft_pct = `FT%`,
    height = `Height (in)`
  ) |>
  mutate(
    ft_pct = as.numeric(gsub("%", "", ft_pct))
  )

# Step 3: Summarize average height and FT% per position
position_summary <- df |>
  group_by(Position) |>
  summarise(
    avg_height = mean(height, na.rm = TRUE),
    avg_ft_pct = mean(ft_pct, na.rm = TRUE),
    count      = n()
  )

# Plot for height distribution by position
df |>
  mutate(Position = reorder(Position, height)) |>
  ggplot(aes(x = Position, y = height, fill = Position)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.size = 2) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = function(x) paste0(x, '"')) +
  labs(
    title    = "Height Distribution by Position",
    subtitle = "Positions ordered by median height",
    x        = "Position",
    y        = "Height (inches)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "grey40"),
    legend.position = "none"
  )

# Table for height summary by position
df |>
  group_by(Position) |>
  summarise(
    `Player Count`       = n(),
    `Avg Height (in)`    = round(mean(height, na.rm = TRUE), 1),
    `Median Height (in)` = round(median(height, na.rm = TRUE), 1),
    `Min Height (in)`    = round(min(height, na.rm = TRUE), 1),
    `Max Height (in)`    = round(max(height, na.rm = TRUE), 1)
  ) |>
  arrange(desc(`Avg Height (in)`)) |>
  kable(
    caption = "Height Summary by Position for Players Who Shot At Least 150 FT's (tallest to shortest)",
    align   = c("l", "c", "c", "c", "c", "c")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE
  )
```

The box plot and summary table together confirm a clear and consistent 
relationship between position and height in the 2025-2026 NBA season. 
From the table, we see that centers are the tallest position with an average height of 82.9 inches, 
while point guards are the shortest at an average of 74.6 inches, a 
difference of over 8 inches. 

Looking at the box plot reveals that the positions 
follow a staircase pattern from PG to C, suggesting that 
height is a strong defining characteristic of positional assignment in 
the NBA. Also, most positions have relatively tight height 
distributions, meaning players at the same position tend to be similar 
in height. This clear correlation between height and position sets up our 
further research into how these attributes affect
the free throw percentage of the players in our dataset.

## Connor

This section examines whether a player’s height is related to their free throw percentage within each position. To evaluate this, we calculate the correlation between height and free throw percentage for each position.

```{r}
library(tidyverse)
library(knitr)

nba_raw <- read_csv("NBAPlayerStats.csv")

nba_clean <- nba_raw |>
  rename(
    position = Position,
    ft_percent = `FT%`,
    height_in = `Height (in)`
  ) |>
  mutate(
    ft_percent = as.numeric(str_remove(ft_percent, "%")),
    height_in = as.numeric(height_in)
  )

position_corr <- nba_clean |>
  group_by(position) |>
  summarise(
    correlation = round(cor(height_in, ft_percent, use = "complete.obs"), 2),
    avg_ft = round(mean(ft_percent, na.rm = TRUE), 1),
    avg_height = round(mean(height_in, na.rm = TRUE), 1),
    n_players = n(),
    .groups = "drop"
  )
```

As shown in @fig-connor, most positions exhibit correlations close to zero, indicating little to no linear relationship between height and free throw percentage within those groups. This suggests that among players who already share a similar role and general body type, being slightly taller or shorter does not meaningfully impact free throw performance.


```{r}
#| label: fig-connor
#| fig-cap: "Correlation between height and free throw percentage by position."
#| fig-alt: "Bar chart showing correlation between height and free throw percentage for each NBA position."
#| fig-pos: "H"

ggplot(position_corr, aes(x = position, y = correlation)) +
  geom_col(fill = "skyblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Position",
    y = "Correlation"
  ) +
  theme_minimal()
```

@tbl-connor provides the corresponding numerical values. While centers show a modest positive correlation and shooting guards show a more noticeable negative correlation, these patterns are inconsistent across positions and likely reflect variation within the sample rather than a strong underlying trend.


```{r}
#| label: tbl-connor
#| tbl-cap: "Correlation between height and free throw percentage by position."

kable(
  position_corr,
  col.names = c("Position", "Correlation", "Avg FT%", "Avg Height", "Players")
)
```

These results suggest that while height is clearly associated with position, it is not a strong predictor of free throw shooting within positions. In other words, once players are grouped by position, differences in height do not explain much of the variation in free throw percentage, indicating that shooting ability is driven more by individual skill than physical size.



## Aidan
Robert confirmed that height varies by position and Connor showed that height has little correlation with accuracy within each position, so I am zooming out and looking at the broader question: 
When we look at each position as a whole, which position tends to shoot free throws better on average?
To explore this, I summarized the data by position and created a bubble chart where each bubble represents one position to show the correlation.

```{r}
#| echo: false

#Tidyverse style guide
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)

# Tidy the data
nba <- read_csv("NBAPlayerStats.csv") |>
  rename(
    ft_pct   = `FT%`,
    height   = `Height (in)`,
    position = Position
  ) |>
  mutate(
    ft_pct = as.numeric(gsub("%", "", ft_pct)),
    height = as.numeric(height)
  )

# Summarize by position
pos_summary <- nba |>
  group_by(position) |>
  summarise(
    avg_height = round(mean(height, na.rm = TRUE), 1),
    avg_ft_pct = round(mean(ft_pct,  na.rm = TRUE), 1),
    n_players  = n(),
    .groups = "drop"
  )
```
@plot-aidan makes the trend pretty easy to see. You can see how there is little difference in the average FT% in guards and small forwards, who also tend to shoot much more accurate than bigs, as they're all fairly even on the Y-axis. The scatter plot helps show a pretty consistent drop off in FT% as height rises, shown by about a 5% drop from small forwards to power forwards, and another 5% drop from power forwards to centers.
We can see that players who are either point guards, shooting guards, or small forwards are on average 79 inches or shorter. From this we can make the observation that players in those positions typically have the ideal attributes to shoot a much more accurate FT% compared to the larger positions in power forwards and centers.  

```{r}
#Plot-aidan — Average height vs. FT% by position
#| label: plot-aidan
#| fig-cap: "Average height vs. average FT% by position. Bubble size shows number of players in each group."

ggplot(pos_summary, aes(x = avg_height, y = avg_ft_pct,
                        size = n_players, color = position)) +
  geom_point(alpha = 0.8) +
  geom_text(aes(label = position), color = "black", size = 4, hjust = 0.5, vjust = 0.5)+  
  scale_size_continuous(range = c(6, 14), name = "# of Players") +
  scale_color_brewer(palette = "Set1", guide = "none") +
  labs(
    title = "Average Height vs. Average Free Throw % by Position",
    x = "Average Height (inches)",
    y = "Average Free Throw %"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold")
  )
``` 
@table-aidan helps viewers clearly see how close the FT% really is between point guards, shooting guards, and small forwards, even given the signifigant average height difference from PG-SF.
Having the actual numbers displayed in addition to the table really helps add emphasis on those three positions shooting within 1.1% of each other on average. 

```{r}
# Table-aidan — Average height vs. FT% by position

#| label: tab-aidan
#| fig-cap: "Provides the corresponding numerical values (Position, Avg Height, Avg FT%, # of Players)."
pos_summary |>
  arrange(desc(avg_ft_pct)) |>
  kable(
    caption = "Average Height and Free Throw Percentage by Position",
    col.names = c("Position", "Avg Height (in)", "Avg Free Throw %", "# Players"),
    align = c("l", "c", "c", "c")
  )
```

# Overall Conclusion
This analysis explored the question of what makes a good NBA free throw shooter by examining the relationship between player height, position, and free throw percentage among players with at least 150 attempts during the 2025-2026 regular season.

Robert's analysis confirmed that height and position are strongly linked, following a clear staircase pattern from point guards (average 74.6 inches) up to centers (average 82.9 inches). This suggests that positional assignment in the NBA is heavily influenced by physical size.

Next, Connor explored how height predicts free throw accuracy within each position. The correlations were consistently weak across all five positions, with the strongest being a modest negative correlation of -0.47 for shooting guards. This indicates that once players are grouped by position, height differences do not meaningfully explain variation in free throw shooting. Instead, individual skill and preparation should be looked at as the primary reason someone is an accurate shooter

Aidan's analysis then zoomed out to compare average free throw percentage across positions as a whole. The results revealed a notable gap between the smaller and larger positions, with point guards, shooting guards, and small forwards shooting above 81% on average, while "bigs" (power forwards and centers) averaged 77.5% and 73.5%. This roughly 5% drop per positional tier suggests that taller players, as a group, tend to be less accurate at the free throw line.

Taken together, the three analyses show that height is not as useful when predicting free throw accuracy among players at the same position. However, at the broader positional level, shorter players tend to shoot free throws more accurately than taller ones position, meaning position itself is a better predictor of free throw accuracy than height alone.


# Author Contributions

- Robert Dang: The first portion of the QMD (marked Robert), GitHub repo setup, the introduction portion, data provenance, FAIR and CARE portions of the QMD

- Connor Box:

- Aidan Wieners:


# AI Usage 
Usage 1: Robert's alt text 05/06/26
GenAI tool used: Anthropic Claude | Claude Sonnet 4.6(claude.ai) 

Prompt: "create an alt text for this data visualization involving nba player height and position df |> mutate(Position = reorder(Position, height)) |> ggplot(aes(x = Position, y = height, fill = Position)) + geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.size = 2) + geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) + scale_fill_brewer(palette = "Set2") + scale_y_continuous(labels = function(x) paste0(x, '"')) + labs( title = "Height Distribution by Position", subtitle = "Positions ordered by median height", x = "Position", y = "Height (inches)" ) + theme_minimal(base_size = 13) + theme( plot.title = element_text(face = "bold"), plot.subtitle = element_text(color = "grey40"), legend.position = "none" ) use the fig-alt feature of R"

Exact Response: "```{r}
#| label: plot-robert
#| fig-cap: "Height Distribution by Position for NBA Players with at Least 150 Free Throw Attempts (2025-2026 Season)"
#| fig-alt: "A box plot showing NBA player height distribution across five positions ordered by median height from shortest to tallest: PG, SG, SF, PF, and C. Each box uses a distinct Set2 palette color with semi-transparent jittered points overlaid. Point guards cluster tightly near 74 inches, shooting guards near 77, small forwards near 79, and power forwards near 80. Centers have the highest median near 83 inches and the widest spread, with outliers exceeding 87 inches. The staircase pattern across positions suggests height is strongly associated with positional assignment in the NBA. 
[code chunk]

The fig-alt chunk option is the Quarto-native way to attach accessibility alt text to a figure. It gets embedded in the rendered HTML output as the alt attribute on the image tag, separate from fig-cap which appears as a visible caption below the plot. If you are rendering to PDF, fig-alt is silently ignored since PDFs handle accessibility differently, but it is still good practice to include it."


# Appendix: Code
Robert's code 
```{r}
#| label: plot-robert
#| fig-cap: "Height Distribution by Position for NBA Players with at Least 150 Free Throw Attempts (2025-2026 Season)"
#| fig-alt: "A box plot showing NBA player height distribution across five positions ordered by median height from shortest to tallest: PG, SG, SF, PF, and C. Each box uses a distinct Set2 palette color with semi-transparent jittered points overlaid. Point guards cluster tightly near 74 inches, shooting guards near 77, small forwards near 79, and power forwards near 80. Centers have the highest median near 83 inches and the widest spread, with outliers exceeding 87 inches. The staircase pattern across positions suggests height is strongly associated with positional assignment in the NBA.
#| eval: false
#| echo: true
#| results: hide

# Tidyverse Style Guide

# Step 1: Load packages
library(ggplot2)
library(dplyr)
library(readr)
library(ggrepel)
library(knitr)
library(kableExtra)

# Step 2: Load and tidy the data
df <- read_csv("NBAPlayerStats.csv") |>
  rename(
    ft_pct = `FT%`,
    height = `Height (in)`
  ) |>
  mutate(
    ft_pct = as.numeric(gsub("%", "", ft_pct))
  )

# Step 3: Summarize average height and FT% per position
position_summary <- df |>
  group_by(Position) |>
  summarise(
    avg_height = mean(height, na.rm = TRUE),
    avg_ft_pct = mean(ft_pct, na.rm = TRUE),
    count      = n()
  )

# Plot for height distribution by position
df |>
  mutate(Position = reorder(Position, height)) |>
  ggplot(aes(x = Position, y = height, fill = Position)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.size = 2) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = function(x) paste0(x, '"')) +
  labs(
    title    = "Height Distribution by Position",
    subtitle = "Positions ordered by median height",
    x        = "Position",
    y        = "Height (inches)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "grey40"),
    legend.position = "none"
  )

# Table for height summary by position
df |>
  group_by(Position) |>
  summarise(
    `Player Count`       = n(),
    `Avg Height (in)`    = round(mean(height, na.rm = TRUE), 1),
    `Median Height (in)` = round(median(height, na.rm = TRUE), 1),
    `Min Height (in)`    = round(min(height, na.rm = TRUE), 1),
    `Max Height (in)`    = round(max(height, na.rm = TRUE), 1)
  ) |>
  arrange(desc(`Avg Height (in)`)) |>
  kable(
    caption = "Height Summary by Position for Players Who Shot At Least 150 FT's (tallest to shortest)",
    align   = c("l", "c", "c", "c", "c", "c")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE
  )
```

Connor's code 
```{r}
#| eval: false
#| echo: true

### Author: Connor Box

library(tidyverse)
library(knitr)

nba_raw <- read_csv("NBAPlayerStats.csv")

nba_clean <- nba_raw |>
  rename(
    position = Position,
    ft_percent = `FT%`,
    height_in = `Height (in)`
  ) |>
  mutate(
    ft_percent = as.numeric(str_remove(ft_percent, "%")),
    height_in = as.numeric(height_in)
  )

position_corr <- nba_clean |>
  group_by(position) |>
  summarise(
    correlation = round(cor(height_in, ft_percent, use = "complete.obs"), 2),
    avg_ft = round(mean(ft_percent, na.rm = TRUE), 1),
    avg_height = round(mean(height_in, na.rm = TRUE), 1),
    n_players = n(),
    .groups = "drop"
  )

ggplot(position_corr, aes(x = position, y = correlation)) +
  geom_col(fill = "skyblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Position",
    y = "Correlation"
  ) +
  theme_minimal()

kable(
  position_corr,
  col.names = c("Position", "Correlation", "Avg FT%", "Avg Height", "Players")
)
```
Aidan's Code
```{r}
#| eval: false
#| echo: true

### Author: Aidan Wieners
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)

# Tidy the data
nba <- read_csv("NBAPlayerStats.csv") |>
  rename(
    ft_pct   = `FT%`,
    height   = `Height (in)`,
    position = Position
  ) |>
  mutate(
    ft_pct = as.numeric(gsub("%", "", ft_pct)),
    height = as.numeric(height)
  )

# Summarize by position
pos_summary <- nba |>
  group_by(position) |>
  summarise(
    avg_height = round(mean(height, na.rm = TRUE), 1),
    avg_ft_pct = round(mean(ft_pct,  na.rm = TRUE), 1),
    n_players  = n(),
    .groups = "drop"
  )

#Plot-aidan — Average height vs. FT% by position
#| label: plot-aidan
#| fig-cap: "Average height vs. average FT% by position. Bubble size shows number of players in each group."

ggplot(pos_summary, aes(x = avg_height, y = avg_ft_pct,
                        size = n_players, color = position)) +
  geom_point(alpha = 0.8) +
  geom_text(aes(label = position), color = "black", size = 4, hjust = 0.5, vjust = 0.5)+  
  scale_size_continuous(range = c(6, 14), name = "# of Players") +
  scale_color_brewer(palette = "Set1", guide = "none") +
  labs(
    title = "Average Height vs. Average Free Throw % by Position",
    x = "Average Height (inches)",
    y = "Average Free Throw %"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold")
  )

# Table-aidan — Average height vs. FT% by position

#| label: tab-aidan
#| fig-cap: "Provides the corresponding numerical values (Position, Avg Height, Avg FT%, # of Players)."
pos_summary |>
  arrange(desc(avg_ft_pct)) |>
  kable(
    caption = "Average Height and Free Throw Percentage by Position",
    col.names = c("Position", "Avg Height (in)", "Avg Free Throw %", "# Players"),
    align = c("l", "c", "c", "c")
  )
```
