---
title: "Stat 184 Free Throw Analysis"
author: "Robert Dang, Connor Box, Aidan Wieners"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
execute:
  echo: false
  warning: false
  message: false
---

# Introduction

This report presents a data analysis and exploration of the question: "What makes a good NBA free throw shooter?" We used player height and position and examined their relationship to free throw percentage among players with at least 150 free throw attempts during the 2025-2026 regular season.

---

# Data Provenance

## Who Collected It
The data was manually collected and compiled by our team. Player 
statistics were gathered by hand from Databallr (databallr.com) and 
entered into an Excel spreadsheet, which was then saved as a CSV file 
for use in this analysis.

## What the Data Is
Each case (row) in the dataset represents one NBA player during the 
2025-2026 regular season. The attributes include player name, team, 
position, physical measurements (height and wingspan in inches), and 
performance statistics such as free throw percentage (FT%) and free 
throw attempts (FTA).

## When It Was Collected
The data was manually recorded on 4/28/2026. The  
statistics reflect the 2025-2026 NBA regular season as displayed 
on Databallr at the time of collection.
---

## Where It Came From
Statistics were transcribed by hand from databallr.com into an Excel 
spreadsheet. The original source of the underlying data is the NBA's 
official game and player records.

## Why It Was Collected
The data was collected to investigate how player height and position relate to free 
throw shooting percentage.

## How It Was Collected
Team members used the features on Databallr to retrieve the necessary statistics. manually looked up each player on Databallr and recorded into an Excel file. This file was then exported as a CSV for use in R. 

---

# FAIR Principles
- **Findable:** The original statistics are publicly viewable on 
  databallr.com. Our compiled dataset is stored in our team's 
  GitHub repository.
- **Accessible:** The CSV file is available in our repo and can be 
  downloaded by anyone with access to it.
- **Interoperable:** The CSV format is compatible with many coding languages
- **Reusable:** Column names are clearly labeled and the data 
  context is well documented in this report, making it reusable 
  for future analysis.

# CARE Principles
This dataset covers NBA players whose statistics are part 
of the public record. No private or sensitive personal information 
was collected beyond what is officially published by the league. 

# Data Cleaning and Feature Creation

Before analysis, the dataset was cleaned and transformed. This included:

- Removing percentage symbols from free throw percentage
- Removing commas from total minutes
- Converting variables to numeric format

These transformations ensure that the data are consistent and suitable for analysis.

---

# Exploratory Data Analysis

## Robert

---

## Connor

This section examines whether a player’s height is related to their free throw percentage within each position. To evaluate this, we calculate the correlation between height and free throw percentage for each position.

```{r}
library(tidyverse)
library(knitr)

nba_raw <- read_csv("NBAPlayerStats.csv")

nba_clean <- nba_raw |>
  rename(
    position = Position,
    ft_percent = `FT%`,
    height_in = `Height (in)`
  ) |>
  mutate(
    ft_percent = as.numeric(str_remove(ft_percent, "%")),
    height_in = as.numeric(height_in)
  )

position_corr <- nba_clean |>
  group_by(position) |>
  summarise(
    correlation = round(cor(height_in, ft_percent, use = "complete.obs"), 2),
    avg_ft = round(mean(ft_percent, na.rm = TRUE), 1),
    avg_height = round(mean(height_in, na.rm = TRUE), 1),
    n_players = n(),
    .groups = "drop"
  )
```

As shown in @fig-connor, most positions exhibit weak relationships between height and free throw percentage.

```{r}
#| label: fig-connor
#| fig-cap: "Correlation between height and free throw percentage by position."
#| fig-alt: "Bar chart showing correlation between height and free throw percentage for each NBA position."
#| fig-pos: "H"

ggplot(position_corr, aes(x = position, y = correlation)) +
  geom_col(fill = "skyblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Position",
    y = "Correlation"
  ) +
  theme_minimal()
```

@tbl-connor provides the corresponding numerical values.

```{r}
#| label: tbl-connor
#| tbl-cap: "Correlation between height and free throw percentage by position."

kable(
  position_corr,
  col.names = c("Position", "Correlation", "Avg FT%", "Avg Height", "Players")
)
```

The results suggest that height is not a strong predictor of free throw performance, as correlations remain small across most positions.

---

## Aidan

---

# Overall Data Story

---

# Conclusion

---

# Open Science and Reproducibility

---

# Author Contributions

- Robert Dang:
- Connor Box:
- Aidan Wieners:


# AI Usage 
---

# Appendix: Code

```{r}

#| eval: false
#| echo: true

### Author: Connor Box

library(tidyverse)
library(knitr)

nba_raw <- read_csv("NBAPlayerStats.csv")

nba_clean <- nba_raw |>
  rename(
    position = Position,
    ft_percent = `FT%`,
    height_in = `Height (in)`
  ) |>
  mutate(
    ft_percent = as.numeric(str_remove(ft_percent, "%")),
    height_in = as.numeric(height_in)
  )

position_corr <- nba_clean |>
  group_by(position) |>
  summarise(
    correlation = round(cor(height_in, ft_percent, use = "complete.obs"), 2),
    avg_ft = round(mean(ft_percent, na.rm = TRUE), 1),
    avg_height = round(mean(height_in, na.rm = TRUE), 1),
    n_players = n(),
    .groups = "drop"
  )

ggplot(position_corr, aes(x = position, y = correlation)) +
  geom_col(fill = "skyblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Position",
    y = "Correlation"
  ) +
  theme_minimal()

kable(
  position_corr,
  col.names = c("Position", "Correlation", "Avg FT%", "Avg Height", "Players")
)
```
